Writing Trash Bin

### 2024-04-03 writingdumpster.qmd
added place for new notes
added notes from @vanderplas2020

>>>>

# New Notes
## @Vanderplas2020

1. Introduction
- The development of graphs and charts has been linked to the development of coordinate systems and abstract representations of data
- the aim of most experimental research in statistical graphics focuses on the initial perception and graph comprehension, very little work has been done to understand the effect of charts and graphs on higher cognitive processes such as learning or analysis
1.1 Design of Statistical Graphs
- improved tech allow charts to be made easily for exploratory purposes
- consider purposes of presentation, entertainmen, analysis. Also considr a continum of utalitarian to artistry as a purpose.
- infographic tells a story (has chart junk) vs a statistical graphic is one of intent (and any visual enhancements should contribute to that aim)
- Points out that the ocean colour and state boundaries in the hurriance map only exist for geographic purposes. The map depicted here has the warnings information along with the cone.
1.2 Statistics Mapping using a Grammar of Graphics
- Systems that attempt to categorize charts based on their geometric representations generally make no effort to include all types of graphics, and they have difficulty accommodating charts that may fall into two or more categories. The classifications of graphics based on the underlying components and their relationships, as in the grammar of graphics developed byWilkinson, are more robust; they also provide an elegant framework for comparing different types of graphical representations separate from the underlying data structure. The former is like treating plots like creatures in a zoo, with a unique name for each, while the latter is analogous to having a phylogeny based on genetic data showing how plots are related.
- Has a visual explanation of the grammar of graphics.
- The grammar of graphics creates visual statistics (and conversely, allows plots to be considered statistics). The data plot can then be treated like other statistics: by imagining what the plot might look like in the absence of any structure, we can use the plot of the actual observed data to test for the likelihood of any perceived structure being significant.
- Using the grammar ofgraphics, it is easy for experimenters to compare different types of charts using the same data, as the underlying structure of the graph remains the same. 
- this approach to transformations and scales allows experimenters to easily test judgments made utilizing different axis transformations and color scales to compare perceptual accuracy

2. Testing Methods
- explicit = partiicpants answer specific questions
- implicitly structured=participants must infer the question of interest from the provided stimuli
2.1 Explicitly structured graphical tests
- Preattentive perceptual effects are those that do not require sustained cognitive attention; they are processed automati- cally within the first 500 ms of viewing a chart or graph. Components processed preattentively include color and shape, as well as some basic information about coarse relationships between individual components
- After the preattentive stage, attention is necessary for subsequent processing. Most of the insights we gain from charts and graphs are due to the cognitive processes that occur after attention is focused on specific aspects of the graph.
2.1.1 Preattentive graph perception
- Preattentively perceived features show a near-constant reaction time over increasing display size, while features that are processed attentively show an increasing reaction time with increased dis- play size
- A primary question in the discussion of preattentive graph perception is whether there are advantages in designing a graph to promote the preattentive perception of features, ideally reducing cognitive load.
- We distinguish between tests that use graphical forms and more primitive tests that use basic geometric elements during the testing process. The results from more primitive experimental designs still apply to the design ofgraphs and charts, but the experimental design does not involve any display of actual data
- Preattentively processed features include shape, angle, size, and texture; however, combinations of preattentive features which represent separate features in the data are processed attentively, with at least one major exception hue and brightness are integrated, that is, that even though they can be separately manipulated, they are still perceived preattentively as a single unit. 
- Experiment on this topic are: HealeyCG,BoothKS, Enns JT. 1996. High-speed visual estimation using preattentive processing; and Callaghan TC. 1984. Dimensional interaction of hue and brightness in preattentive field segregation. Percept.
2.1.2 Attention medicated testing methods
2.1.2.1 Direct Observation: numerical estimation, speed, error rates
- simplest ways to test the utility ofa graph is to verify that information can be accurately read from it
- the participants are provided with a chart and asked to estimate some quantity or answer a predefined question using the information provided in the chart
- It is important to ensure that the specific charts and questions used are aligned; studies are commonly critiqued on the basis that the charts or the questions were not appropriate for the task
- Not always perceptual ranking. The point vs line example in figure 8 shows there is some cases where perceptual tasks do not perfectly differentiate? Or is it that the PDF you highlight is of importance.
- open-ended estimation tasks elicit certain well-known biases such as the tendency to round to multiples of 5 or 10
- To approach situations with more complicated graphics, or charts that are known to induce perceptual biases in the participants, consider using eye tracking to measure attention and motivation, or the use of verbal descriptions to assess more complicated graphs (since they facilitate specific parts of the perceptual process)
2.1.2.2 Psychophysics and signal detection theory
- Psychophysical experimental design is focused on whether an effect is detectable and whether the magnitude of the effect can be accurately estimated
- Common methods, such as the method of constant stimuli and the method of adjustment, involve repeatedly presenting a participant with charts and asking them to evaluate the chart on the basis of a particular question of interest
- In the method of adjustment, this is done with the control of the participant, who adjusts the stimuli interactively until the effect is just barely noticeable; in the method of constant stimuli, the effect size changes randomly from trial to trial to reduce continuity effects.
- Psychophysics methods also seem to be relatively common in studies of map perception, particularly when the goal is to estimate the amount of exaggeration or other corrective distortion necessary for realistic perception of the map (for
2.1.2.3 Thinking aloud
- ask participants to talk through their thoughts as they read and use a graph in a realistic setting
- allows experimenters to examine the use ofcom- plex graphics in the wild, or at least in situations that are less artificial than the paradigms allowed by numerical estimation and psychophysics methods
- allow researchers to attempt to measure insight, and reasoning in complex situations such as experimental design, decision-making, or the process of weather fore- casting
- While the data that result from the think-aloud protocol are typically more qualitative and less quantitative than results produced using other methods, they provide significant additional insight into the underlying cognitive processes affecting visualization, which cannot be obtained through other means
2.1.2.4 Eye tracking
- eye tracking facilitates insight into the process of visual attention, providing data on the approximate spatial location of visual focus
- Eye tracking allows researchers to determine that viewers spend relatively little time examining the axes in scatter plots, but significant amounts of time examining the axes in parallel coordinates plots, suggesting that the process of reading these two chart types is fundamentally different
- Can also be used to identify features that provide useful information during the graph reading process for several different types of charts
2.1.2.5 Combination Experiemnts
2.2 Implicit Graphical Tests Using Visual Inference
- In an implicit graphical test, the user must identify both the purpose and function of the plot and use that information to evaluate the plot as shown
- Explicit tests are typically conducted on plots that have been created to showcase specific structure in the data in order to present results; in contrast, implicit tests are designed to inform exploratory data analysis and the iterative model diagnostic process. 
- e.g. lineup plot
- The advantage to implicit testing is that lineups do not require a specification of a feature of interest in the testing framework
- Much ofthe historical research of comparing different types of charts has been criticized because the specific question phrasing does not provide readily generalizable results; the lineup protocol removes this obstacle by charging the user with the task of identifying the most different looking plot and thereby selecting the feature with the visually most salient difference compared with the other plots. 
- The real power ofthe lineup protocol is that when combined with the grammar of graphics, we can hold the underlying data and summary statistics constant, isolating the effect ofdifferent plot types, coordinate transformations, and aesthetic mappings on our ability to detect effects in the data.
- It can be difficult to specify the null data-generating model in a way that adequately mimics the data plots, which suggests that visually, we are able to identify many more features than those typically tested using standard quantitative hypothesis tests. This implicit testing of many different hypotheses does make null distribution specification challenging, but it also highlights the power ofvisual cognition to detect subtle differences in data.

3 Current Best Graphical Practice
- no best plot can lead to a rather fragmented approach when describing best practice, and so in order to avoid this, we examine graphical practice using the principle of “first, do no harm” from the Hippocratic oath.
3.1 Cognitive Principles
- Take advantage of human visual system and allows us to make charts that require less cognitive effort to read
3.1.1 Proximity
- objects or shapes that are close to one another appear to form groups. For plot design, proximity is used to place items to compare close together, and less important comparisons further apart.
3.1.2 Similarity
- The gestalt principle of similarity suggests that we group things that have similar appearances and exclude objects with different appearances. 
. Vanderplas & Hofmann (2017) showed that the addition of color and shape to a scatter plot increases the likelihood that individuals will perceive clustered groups of points.
- the coloring of bars allows us to easily see that the sim- ilarly colored rectangles represent the same group of people, even though the bars are separated by facets and other groups.
3.1.3 Common Region
- The gestalt principle of common region suggests that elements contained within a common region belong together.
- confidence bands and bounding ellipses also activate this gestalt principle by grouping points within the boundaries together (Vanderplas & Hofmann 2017), highlighting the presence of outliers that do not belong to the main group.
3.1.4 Common Fate
- The gestalt principle of common fate describes the tendency to group objects that are moving together in the same direction and at the same speed together
- Common fate is certainly active in animated plots that use fading or transitions over time, but even in static plots, continuity can be activated when multiple time series plots are shown together (common fate creates bias in identifying correlation)
- Strong negative association is not easily detected from overlaid line plots, but it is easily seen in a scatter plot (makes me think of that HOPS plot experiment)
3.1.5 Working Memory
- Typically, working memory is limited to approximately seven (plus or minus two) items, or chunks. In practice, this means that categorical scales with more than seven categories decrease readability, increase comprehension time, and require signifi- cant attentional resources, because it is not possible to hold the legend mapping in working memory.
3.1.6 Change Blindness
- change blindness, affects both static and interactive plots. In static plots, it can be difficult to compare between dif- ferent small multiples or facets because the contents of the plots are not reliably represented in workingmemorywhen switching attention between them. In animated plots, it is important to use transition effects to connect successive frames of the animation: This reduces change blindness and also activates the gestalt principle of common fate, allowing us to quickly identify groups of objects that are transitioning in the same direction
3.1.7 Ease of Comparison
- Much of the psychophysics research on statistical charts exam- ines the accuracy of comparisons and quantitative evaluations made during the process of under- standing a plot.
-This ranking ofcognitive tasks provides some consistent guidance for chart design: If the same data can be represented in a way that allows the user to make a comparison more accurately (based on the hierarchy), then that design is preferable.
- While the hierarchy of graphical comparisons provides some guidance, there are other design choices that can be informed by experimental research in a less systematic way.
3.2 Chart Design
- The use of redundant aesthetics that activate the same gestalt principles (such as color and shape in a scatter plot, which both activate similarity) results in higher identification ofcorresponding data features. 
- In addition, dual encoding increases the accessibility ofa chart to individuals who have impaired color vision or perceptual processing (e.g., dyslexia, dysgraphia).
3.2.1 Colour
- Cognitive load can also be reduced by selecting colors with cultural associations that match the data display, such as the use of blue for men and red (or pink) for women, or the use ofblue for cold temperatures and red/orange for warm temperatures.
- It is also important to consider the human perceptual system, which does not perceive hues uniformly: We can distinguish more shades of green than any other hue, and fewer shades of yellow, so green univariate color schemes will provide finer discriminability than other colors be- cause the human perceptual system evolved to work in the natural world, where shades of green are plentiful.
- the distance between points in a given color space may not be the same as the distance between points in perceptual space
- Color-blindness is common enough that it is reasonable to expect that any given chart used in a presentation or publication will be read by someone with a color perception deficiency. The use of dual encoding allows color-blind individuals to more readily read graphics that utilize color, and as hue and lightness can be varied separately, it is possible to use dual encoding without adding another aesthetic.

4. Open Questions and Future Research
- Another open question is the acceptance of results—much historical research on best practices exists, but how much of it is being put into practice? One

## @Tierney2023
